Quantifying the importance of an inference model in Bayesian phylogenetics

### 2020-07-14 pirouette_article.tex

\new{
	Many other diversification models (and their associated likelihood algorithms) 
	have been developed, e.g., models in which diversification 
	is time-dependent [\cite{nee1994reconstructed,rabosky2008explosive}],
	or diversity-dependent [\cite{etienne2012diversity}],
	or where diversification rates change for specific lineages 
	and their descendants [\cite{etienne2012conceptual, rabosky2014automatic, alfaro2009nine, laudanno2020sls}].
	Other models treat speciation as a process that takes 
	time [\cite{rosindell2010protracted, etienne2012prolonging, lambert2015reconstructed}],
	or where diversification rates
	depends on one or more traits 
	[\cite{maddison2007estimating, fitzjohn2012diversitree, herrera2018detecting}].
}

@@

@@

Here we introduce a method to quantify the \new{impact} 
of a novel tree prior,
\new{
  i.e., a tree model, for which we can simulate phylogenies, but not yet
  calculate their likelihoods.
}
\new{
  This new method simultaneously assesses the substitution, clock and tree
  models and thus is a hierarchical model.
  We learn from \cite{duchene2015evaluating} that 
  such a method may be sensitive to other aspects of the hierarchical model.
}
The method starts with a phylogeny generated by the new model. 
Next, nucleotide sequences are simulated that follow the evolutionary 
history of the given phylogeny. 
Then, using BEAST2's built-in tree priors,
a Bayesian posterior distribution of phylogenies is inferred. 
We then compare the inferred with the original phylogenies. 
How to properly perform this comparison forms the heart of our method.
Only new diversification models that result 
in a large discrepancy between inferred and simulated phylogenies 
will be worth the effort and computational burden to implement 
a species tree prior for in a Bayesian framework.

@@

\begin{figure}
  \centering
  \includegraphics[width = 0.99\textwidth]{workflow4.png}
  \caption{
    \texttt{pirouette} pipeline.
    The pipeline starts from a phylogeny (1a) simulated by the 
    generative tree model 
    $\mathit{p_{G}}$.
    The phylogeny is converted to an alignment (2a) using the generative 
    alignment model 
    $\mathit{A} = (\mathit{c_{G}}, \mathit{s_{G}})$, composed of a clock model 
    and a site model. 
    The user defines one or more experiments.
    For each candidate experiment $\mathit{X_{i}}$ 
    (a combination of inference model $\mathit{I_{i}}$ and condition $\mathit{C_{i}}$),
    if its condition $\mathit{C_{i}}$ is 
    satisfied (which can depend on the alignment), 
    the corresponding inference model $\mathit{I} = \mathit{I_{i}}$ is selected
    to be used in the next step.
    The inference models (3a) of the selected experiments use the alignment (2a) 
    to each create a Bayesian posterior of (parameter estimates and) 
    phylogenies (4a). 
    Each of the posteriors' trees is compared to the true phylogeny (1a) 
    using the error measure $\mathit{E}$, 
    resulting in an error distribution (5a). 
    Optionally, for each selected inference model a twin pipeline can be run.
    A twin phylogeny (1b) can be generated from the original 
    phylogeny (1a) using the twin tree model $\mathit{p_{t}}$, 
    selected among standard diversification models; 
    the default option is the standard BD model, 
    with parameters estimated from the original phylogeny.
    A twin alignment (2b) is then simulated from the twin phylogeny 
    using clock model $\mathit{c_{G}}$ and site model $\mathit{s_{G}}$ 
    imported from the generative model (the model that has
    actually generated the phylogeny, or is assumed to have done so). 
    The twin pipeline follows the procedure of the main pipeline, 
    resulting in a twin error distribution (5b).
  }
  \label{fig:pipeline}
\end{figure}

@@

The pipeline is visualized in Fig.~\ref{fig:pipeline}. 
There is also the option to generate a 'twin tree', 
that goes through the same pipeline (see section \ref{subsec:twinning}).

@@

Similar to lab experiments, \verb;pirouette; allows to perform
two types of control measurements. The first type of control
is called 'twinning', which results in an error distribution
that is the baseline error of the inference pipeline (see
supplementary materials, section \ref{subsec:twinning} for more details).
The second type of control is the use of candidate models,
which results in an error distribution
for a generative model that is determined to be the best fit to
the tree (see supplementary materials, 
section \ref{subsec:candidates} for more details).
Additionally, multiple \verb;pirouette; runs are needed
to reduce the influence of stochasticity (see supplementary materials, 
section \ref{subsec:stochasticity} for more details).

@@

The error distribution shown in Figure \ref{fig:example_30}
is produced, which uses the nLTT 
statistic [\cite{janzen2015approximate}] to compare phylogenies (see 
section \label{subsec:nltt} for details regarding the nLTT statistic
and its caveats).

@@

Interpreting the results of \verb;pirouette; is up to the user; 
\verb;pirouette; does not answer the question 
whether the inference error is too large to trust the inferred tree. The user is encouraged to use different statistics to measure the error. The nLTT statistic is
a promising starting point, as it can compare any two trees and 
results in an error distribution of known range, but one may also explore other statistics.
In principle, \verb;pirouette; allows for this, but in our example we used a diversification model (DD) that only deviates from the Yule and BD models in the temporal branching pattern, not in the topology.
\new{
  For models that make different predictions on topology, 
  the twinning process should be modified accordingly.
}
\new{
	As noted in the introduction, Duch\^{e}ne and 
	colleagues [\cite{duchene2018phylodynamic}]
	also developed a method to assess the adequacy of a tree model
	on empirical trees. They simulate trees from the posterior distribution of 
	the parameters and then compare this to the originally inferred tree using 
	tree statistics, to determine whether the assumed tree model in inference 
	indeed generates the tree as inferred. This is useful if these trees match, 
	but when they do not, this does not mean that the inferred tree is incorrect; 
	if sufficient data is available the species tree prior may not be important, 
	and hence the inference may be adequate 
	even though the assumed species tree prior is not. 
	In short, the approach is applied to empirical trees and 
	compares the posterior and prior distribution of trees (with the latter 
	generated with the posterior parameters!).
}
\verb;pirouette;
\new{
	aims to expose when assuming standard priors for the 
	species tree are a mis- or underparameterization. 
	Our approach applies to simulated trees and compares 
	the posterior distributions of trees generated with 
	a standard and non-standard model, but inferred with a standard one. 
	The two methods therefore complement one another.
} However, we note that the \verb;pirouette; pipeline is not restricted 
to exploring the effects of a new species tree model. 
The pipeline can also be used to explore the effects of non-standard 
clock or site models, such as relaxed clock models with a non-standard 
distribution, correlated substitutions on sister lineages, or elevated 
substitutions rates during speciation events. 
It is, however, beyond the scope of this paper to discuss all these options 
in more detail.

>>>>

\new{
	Many other diversification models (and their associated likelihood algorithms) 
	have been developed, e.g., models in which diversification 
	is time-dependent [\cite{nee1994reconstructed,rabosky2008explosive}],
	or diversity-dependent [\cite{etienne2012diversity}],
	or where diversification rates change for specific lineages 
	and their descendants [\cite{etienne2012conceptual, rabosky2014automatic, alfaro2009nine, laudanno2020detecting}].
	Other models treat speciation as a process that takes 
	time [\cite{rosindell2010protracted, etienne2012prolonging, lambert2015reconstructed}],
	or where diversification rates
	depends on one or more traits 
	[\cite{maddison2007estimating, fitzjohn2012diversitree, herrera2018detecting}].
}

@@

\new{
Similarly,
Ritchie and colleagues [\cite{Ritchie2017impact}] showed that inference was accurate when birth-death or skyline coalescent priors were used, but they simulated their trees with a Yule process only, as their focus was not so much on the diversification process but on the influence of inter- and intraspecific sampling.
}

@@

Here we introduce a method to quantify the \new{impact} 
of a novel tree prior,
\new{
  i.e., a tree model, for which we can simulate phylogenies, but not yet
  calculate their likelihoods.
}
\new{
  This new method simultaneously assesses the substitution, clock and tree
  models [\cite{duchene2015evaluating}].
}
The method starts with a phylogeny generated by the new model. 
Next, nucleotide sequences are simulated that follow the evolutionary 
history of the given phylogeny. 
Then, using BEAST2's built-in tree priors,
a Bayesian posterior distribution of phylogenies is inferred. 
We then compare the inferred with the original phylogenies. 
How to properly perform this comparison forms the heart of our method.
Only new diversification models that result 
in a large discrepancy between inferred and simulated phylogenies 
will be worth the effort and computational burden to implement 
a species tree prior for in a Bayesian framework.

@@

\begin{figure}
  \centering
  \includegraphics[width = 0.97\textwidth]{workflow4.png}
  \caption{
    \texttt{pirouette} pipeline.
    The pipeline starts from a phylogeny (1a) simulated by the 
    generative tree model 
    $\mathit{p_{G}}$.
    The phylogeny is converted to an alignment (2a) using the generative 
    alignment model 
    $\mathit{A} = (\mathit{c_{G}}, \mathit{s_{G}})$, composed of a clock model 
    and a site model. 
    The user defines one or more experiments.
    For each candidate experiment $\mathit{X_{i}}$ 
    (a combination of inference model $\mathit{I_{i}}$ and condition $\mathit{C_{i}}$),
    if its condition $\mathit{C_{i}}$ is 
    satisfied (which can depend on the alignment), 
    the corresponding inference model $\mathit{I} = \mathit{I_{i}}$ is selected
    to be used in the next step.
    The inference models (3a) of the selected experiments use the alignment (2a) 
    to each create a Bayesian posterior of (parameter estimates and) 
    phylogenies (4a). 
    Each of the posteriors' trees is compared to the true phylogeny (1a) 
    using the error measure $\mathit{E}$, 
    resulting in an error distribution (5a). 
    Optionally, for each selected inference model a twin pipeline can be run.
    A twin phylogeny (1b) can be generated from the original 
    phylogeny (1a) using the twin tree model $\mathit{p_{t}}$, 
    selected among standard diversification models; 
    the default option is the standard BD model, 
    with parameters estimated from the original phylogeny.
    A twin alignment (2b) is then simulated from the twin phylogeny 
    using clock model $\mathit{c_{G}}$ and site model $\mathit{s_{G}}$ 
    imported from the generative model (the model that has
    actually generated the phylogeny, or is assumed to have done so). 
    The twin pipeline follows the procedure of the main pipeline, 
    resulting in a twin error distribution (5b).
  }
  \label{fig:pipeline}
\end{figure}

@@

The pipeline is visualized in Fig.~\ref{fig:pipeline}. 
There is also the option to generate a 'twin tree', 
that goes through the same pipeline (see supplementary subsection \ref{subsec:twinning}).

@@

Similar to lab experiments, \verb;pirouette; allows to perform
two types of control measurements. The first type of control
is called 'twinning', which results in an error distribution
that is the baseline error of the inference pipeline (see
supplementary materials, subsection \ref{subsec:twinning} for more details).
The second type of control is the use of candidate models,
which results in an error distribution
for a generative model that is determined to be the best fit to
the tree (see supplementary materials, 
section \ref{subsec:candidates} for more details).
Additionally, multiple \verb;pirouette; runs are needed
to reduce the influence of stochasticity (see supplementary materials, 
section \ref{subsec:stochasticity} for more details).

@@

The error distribution shown in Figure \ref{fig:example_30}
is produced, which uses the nLTT 
statistic [\cite{janzen2015approximate}] to compare phylogenies (see 
section \ref{subsec:nltt} for details regarding the nLTT statistic
and its caveats).

@@

Interpreting the results of \verb;pirouette; is up to the user; 
\verb;pirouette; does not answer the question 
whether the inference error is too large to trust the inferred tree. The user is encouraged to use different statistics to measure the error. The nLTT statistic is
a promising starting point, as it can compare any two trees and 
results in an error distribution of known range, but one may also explore other statistics.
In principle, \verb;pirouette; allows for this, but in our example we used a diversification model (DD) that only deviates from the Yule and BD models in the temporal branching pattern, not in the topology.
\new{
  For models that make different predictions on topology, 
  the twinning process should be modified accordingly.
}
\new{
	As noted in the introduction, Duch\^{e}ne and 
	colleagues [\cite{duchene2018phylodynamic}]
	also developed a method to assess the adequacy of a tree model
	on empirical trees. They simulate trees from the posterior distribution of 
	the parameters and then compare this to the originally inferred tree using 
	tree statistics, to determine whether the assumed tree model in inference 
	indeed generates the tree as inferred. This is useful if these trees match, 
	but when they do not, this does not mean that the inferred tree is incorrect; 
	if sufficient data is available the species tree prior may not be important, 
	and hence the inference may be adequate 
	even though the assumed species tree prior is not. 
	In short, the approach is applied to empirical trees and 
	compares the posterior and prior distribution of trees (with the latter 
	generated with the posterior parameters!).
    By contrast,}
    \verb;pirouette;
	\new{aims to expose when assuming standard priors for the 
	species tree are a mis- or underparameterization. 
	Hence, our approach applies to simulated trees and compares 
	the posterior distributions of trees generated with 
	a standard and non-standard model, but inferred with a standard one. 
	The two methods therefore complement one another.
}
However, we note that the \verb;pirouette; pipeline is not restricted 
to exploring the effects of a new species tree model. 
The pipeline can also be used to explore the effects of non-standard 
clock or site models, such as relaxed clock models with a non-standard 
distribution, correlated substitutions on sister lineages, or elevated 
substitutions rates during speciation events. 
It is, however, beyond the scope of this paper to discuss all these options 
in more detail.
